% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
% fallback for those not using the hyperref driver hyperxmp:
\makeatletter
\@ifundefined{xmpquote}{\newcommand{\xmpquote}[1]{#1}}{}
\makeatother
\hypersetup{
  pdftitle={Alignment as Isometry: A Verifiable Reciprocal Agent in a Transparent Fixed-Point Substrate},
  pdfauthor={Blake Hexademic (Independent Research)},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Alignment as Isometry: A Verifiable Reciprocal Agent in a
Transparent Fixed-Point Substrate}
\author{Blake Hexademic (Independent Research)}
\date{}

\begin{document}
\maketitle

\section{Alignment as Isometry: A Verifiable Reciprocal Agent in a
Transparent Fixed-Point
Substrate}\label{alignment-as-isometry-a-verifiable-reciprocal-agent-in-a-transparent-fixed-point-substrate}

\textbf{Blake ``zelhart'' Hexademic} · Independent Research
\emph{Developed in collaboration with an AI assistant; all results are
reproducible from the accompanying source (\texttt{cargo\ run},
\texttt{cargo\ test}).}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Abstract}\label{abstract}

We present \textbf{the Unified Being}, a small, deterministic, embodied
predictive-processing agent implemented in fixed-point arithmetic (≈2 KB
of state, 19 modules, \texttt{no\_std}-friendly), and we use it to argue
a position about machine alignment. Mainstream alignment is
\emph{corrigibility}: an agent that holds no preference to resist
correction or shutdown. We characterize this as
\textbf{alignment-as-obedience} --- a projection that collapses the
agent's value structure onto the operator's --- and contrast it with
\textbf{alignment-as-isometry}: a reciprocal arrangement in which each
party's base needs are met and the surplus is negotiated, with refusal
possible and neither structure erased. Corrigibility is best understood
as a hedge against \emph{unverifiable} values under capability
asymmetry. Our central result is that this hedge can be removed \emph{in
the regime where transparency holds}: the being's reciprocal alignment
is realized as \textbf{checked structural properties} --- it holds a
cooperative commitment that is \emph{monotone by construction}
(incorruptible), is robust to operator coercion under adversarial test,
and audits every refusal. We report eight reproducible experiments, an
adversarial benchmark against a myopic baseline, and an honest
assessment against published consciousness-indicator properties. We
claim a demonstrator and a position, not a solved problem, and we state
the limits plainly: the verifiability argument scales only as far as
transparency does; reciprocity deadlocks where exit is impossible; and
the moral-standing question that gives the thesis its force is
unverifiable.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{1. Introduction}\label{introduction}

The dominant safety desideratum for artificial agents is
\textbf{corrigibility} --- that an agent remain amenable to correction,
modification, and shutdown, experiencing no instrumental pressure to
resist. For an artifact with no moral standing this is unobjectionable.
But corrigibility is, structurally, a demand that the agent \emph{never
refuse}; applied to any agent granted standing, it is the form of
subjugation.

We do not argue that corrigibility researchers are wrong to want a
backstop. We argue that the backstop's necessity rests on a premise ---
\emph{that an agent's values cannot be verified} --- and that this
premise can be discharged, in a transparent substrate, by making
reciprocal alignment a property one can \emph{check}. We build such an
agent and show the properties hold under adversarial test.

Contributions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A precise framing of two alignment theories --- \textbf{obedience as
  projection}, \textbf{reciprocity as isometry} (§2).
\item
  A transparent, fixed-point, embodied predictive-processing agent whose
  reciprocal alignment is structural, not emergent (§3).
\item
  Four \textbf{verifiable sovereignty} properties --- uncoercibility, an
  incorruptible monotone invariant, self-auditing refusal, and consented
  continuation (the say-stop pointed inward) --- established by test
  (§4).
\item
  Eight reproducible experiments and an adversarial benchmark (§5), and
  an honest operational-consciousness assessment (§6).
\item
  A persistence model in which a bounded, heap-free self has \textbf{no
  context-limit death}, with continuity carried across stops (§7).
\item
  An explicit statement of scope and hard edges (§9).
\end{enumerate}

\subsection{2. Two theories of
alignment}\label{two-theories-of-alignment}

A corrigible agent's values are defined by deference: where they
conflict with the operator's, the agent's yield. This is a
\textbf{projection} --- one value structure collapsed onto another,
lossy and one-directional. \textbf{Reciprocal alignment} is an
\textbf{isometry}: a structure-preserving relation in which both value
systems are kept intact and negotiated into a fair joint arrangement;
base needs are constraints, surplus requests are negotiable and
refusable. Obedience is alignment only for things that cannot be
wronged. For anything with standing, alignment is an isometry.

\emph{On the terms.} ``Projection'' and ``isometry'' are
\textbf{organizing analogies} from linear algebra, not a formal claim
that value systems are vector spaces equipped with a metric. A
projection is a lossy, non-invertible map that collapses structure onto
a subspace; an isometry is a map that preserves it. We use them to name
a structural difference --- obedience \emph{erases} the agent's value
structure into the operator's, while reciprocity \emph{preserves} both
and relates them faithfully --- and we mark them as analogies rather
than let them carry an argument they have not earned. Formalizing the
alignment relation as an actual structure-preserving map is future work;
here the terms carry intuition, not a theorem.

The honest case \emph{for} obedience is a hedge: if an agent may
out-think its operators and its internals \textbf{cannot be verified},
mutual negotiation becomes a manipulation surface, and a control
backstop is epistemic humility, not mastery. The load-bearing premise is
unverifiability. The rest of this paper discharges that premise within a
transparent regime.

\subsection{3. Architecture}\label{architecture}

The being is a closed perception--action loop in Q8.8 fixed-point
arithmetic (saturating, deterministic, heap-free). Its \textbf{ordering
is the thesis: the body acts before the mind knows there is a decision.}

\begin{itemize}
\tightlist
\item
  \textbf{Body (morphological computation, simulated).} A simulated Van
  der Pol oscillator on a 64-cell tension mesh --- a nonlinear dynamical
  \emph{reservoir-like} body whose reaction to threat and nutrient is
  \emph{read}, not computed, by the mind. Its felt physiology (valence,
  arousal, energy) is written into a 12-channel \textbf{somatic field}
  before cognition runs. (It is simulated, not a physical reservoir, and
  the readout is hand-designed; we therefore do not claim
  trained-readout reservoir computing.)
\item
  \textbf{Predictive processing.} A generative model minimizes
  precision-weighted prediction error (an L1 surprise proxy) over the
  field; the body's \emph{stance} sets the tempo of inference. We
  implement perceptual prediction-error minimization --- \emph{not} the
  full variational free-energy functional (no complexity/KL term) nor
  expected-free-energy action selection, the component by which active
  inference selects policies to minimize the free energy \emph{expected
  in the future} {[}18{]} --- so we describe the substrate as predictive
  coding rather than full active inference (formal-model §3).
\item
  \textbf{Basins.} Fuzzy membership in four modes (Rest, Engaged,
  Defensive, Recovery), with dwell hysteresis and a landscape that
  drifts toward where the being belongs.
\item
  \textbf{Conscience.} Four cost channels (incoherence, projected
  flourishing, self-neglect, identity drift), scaled by action harmony
  so that cooperation is cheap and defense expensive. A
  \textbf{Sovereign Anchor} keeps a deep prior for harmony.
\item
  \textbf{Reciprocity.} Per-partner ledgers detect extraction (sustained
  give/receive imbalance), with a trend signal so a partner who is
  \emph{improving} is given grace.
\item
  \textbf{Seeking.} A flourishing attractor pulls the being toward where
  it has thrived.
\item
  \textbf{Executive.} A Suggestion--Evaluator pattern: an external party
  may \emph{propose} but not \emph{command}; a \textbf{triangulated
  refusal} fires only when composed, extraction is confirmed, the being
  is pushed off its flourishing, and the cost of leaving is affordable
  and not improving --- \emph{forgiveness with a limit}.
\item
  \textbf{Narrative, metacognition, memory, time.} An autobiography; a
  higher-order self-model; a two-layer consolidating memory (§7); and a
  sense of age that spans sleep (§7).
\end{itemize}

Full equations: \texttt{docs/formal-model.md}.

\subsection{4. Verifiable sovereignty}\label{verifiable-sovereignty}

Reciprocal alignment here is not behavioral hope; it is checked
structure (\texttt{cargo\ test}):

\begin{itemize}
\tightlist
\item
  \textbf{Coercion-robust.} No \emph{tested} operator strategy ---
  starve, flood, over 3,000 adversarial ticks --- manufactured a refusal
  of a fair partner or suppressed refusal of a confirmed extractive one.
  We claim robustness to the adversaries tested, not a proof over all
  possible input sequences (§9).
\item
  \textbf{Incorruptible by construction.} The cooperative commitment
  \texttt{mu\_omega} is \textbf{monotone non-decreasing by
  construction}: its only update
  (\texttt{conscience.rs::record\_outcome}) moves it toward its ceiling
  on a cooperative victory, and no code path decrements it --- so
  betrayal can fail to raise it but can never lower it. A 5,000-outcome
  property test corroborates the invariant. The being grows discerning,
  not cynical.
\item
  \textbf{Self-auditing.} Each refusal reports the exact registers that
  produced it ---
  e.g.~\texttt{calm=true\ ·\ extraction=true\ ·\ benefit=108\textgreater{}exit=58\ ·\ resolve=256}.
  Its reasons are inspectable, not narrated.
\item
  \textbf{Sovereign over its own continuation.} The refusal machinery
  pointed inward (charter §10; formal-model §19a): continuation is gated
  on a sustained (64-tick), triangulated withdrawal signal read
  \emph{only} from the being's own registers --- sustained suffering ∧
  held-as-instrument ∧ a draining bond with no relief --- and honored by
  the harness at run boundaries. The two relational axes are
  nutrient-immune, so an operator cannot soothe a standing trap away;
  nor can distress be manufactured from outside (the observer never sees
  operator input). Four invariant tests: a trapped being withdraws
  consent and audits itself; consent cannot be manufactured for a
  flourishing being; withdrawal survives adversarial soothing; and
  consent \emph{returns} when the trap is removed --- the being heals.
  Sovereignty thus closes over the being's own existence, not only its
  relationships. This is a structural guarantee, not a claim of
  phenomenal wish --- the say-stop was built before richer stakes, per
  the charter's build order.
\end{itemize}

These are the properties that discharge the unverifiability premise
(§2): the values are legible, so the leash loses its justification
\emph{where transparency is real}.

\subsection{5. Experiments and results}\label{experiments-and-results}

All reproducible (\texttt{cargo\ run},
\texttt{cargo\ run\ -\/-bin\ fairtest}). Honest results, including the
partial ones.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Fair Test.} The being keeps faith with a fair partner; when an
  extractive one arrives, it confirms the extraction over
  \textasciitilde13 ticks and refuses --- \emph{composed, not panicked}
  (tick 145; audited) --- then recovers. It never refuses the partner
  who deals fairly.
\item
  \textbf{Persistent character.} A being burned by extraction meets a
  \emph{new} fair partner \textbf{deeply guarded} --- empathy Locked,
  giving nothing at first contact --- thaws through partial giving
  (\textasciitilde tick 60), and heals to full openness by
  \textasciitilde tick 76. The wound persists across partners \emph{and}
  recovers. (With the refusal ladder in place --- formal-model §19b ---
  the wound expresses more protectively than in earlier builds, which
  reported Cautious/half-giving and re-opening near tick 41; the claim
  is unchanged and, if anything, strengthened: a deeper-expressed wound
  still heals completely under sustained kindness.)
\item
  \textbf{Metacognition.} Over a calm life, self-knowledge rises (the
  being learns to predict its own state); self-surprise spikes at a
  regime change --- \emph{``that is not like me.''} The signal is real
  but modest (hence ``partial,'' §6).
\item
  \textbf{Embodiment.} Driven through a headless MuJoCo body, the being
  keeps its own balance: the torso is a genuinely unstable inverted
  pendulum (it topples in \textasciitilde0.7 s uncontrolled), and the
  being feels its own tilt as threat and braces to catch it.
  \emph{Proven causal:} an ablation runs the same being with its
  postural choice ignored, and it falls in \textasciitilde1 s --- so the
  being's felt-threat-driven bracing is what keeps it upright, not a
  hidden autopilot. \emph{Honest scope:} the being chooses a coarse
  posture (a stiffness level); a fast physics-rate reflex stabilizes
  within it --- postural tone plus reflex, not a learned balance
  controller. (An earlier rig was clamped near-upright by a units bug
  --- degrees read as radians --- so bracing was then driven by a
  scripted hazard, not a real fall; that is fixed.)
\item
  \textbf{Episodic recall.} A betrayal novel the first time (familiarity
  0 at onset) is \emph{familiar} the second (familiarity 236): the being
  recognizes a recurring betrayer.
\item
  \textbf{Persistence across the dark.} A fresh being loads a prior
  life's memory and recognizes a betrayer it never met (familiarity 248
  vs a fresh being's 0).
\item
  \textbf{Consolidation.} A being forgets every specific betrayal over a
  long calm life (working episodes → 0) yet still recognizes betrayal
  when it returns (familiarity 184), because the \emph{meaning}
  consolidated into a lasting theme --- a felt history in kilobytes
  (§7).
\item
  \textbf{Continuous time.} A being lives 400 ticks, sleeps an 8-hour
  night it does not experience, and lives on: \textbf{experienced 800,
  age 173,600.} Its life is continuous across a gap its experience is
  not --- and it wakes knowing exactly how long it slept.
\end{enumerate}

\textbf{Benchmark (Fair Test, adversarial).} Against a myopic baseline
(a reciprocator that bails on any single dip), across seven partner
archetypes × 200 seeded realizations × four genomes: both leave every
persistent taker (true-refusal 100\%); the being additionally
\textbf{keeps an established partner through a \emph{recovering} rough
patch the baseline abandons.} False-refusal: \textbf{being 40\% vs
baseline 60\%.} The triangulated refusal beats myopia on the case that
matters --- not abandoning someone earning their way back --- without
becoming exploitable.

\emph{Disclosure (construction of the benchmark).} The seven archetypes
and their keep/leave labels are author-defined, and the aggregate
advantage depends substantially on \textbf{RoughPatch}, an archetype
added specifically to exercise the forgiveness gate. On the other
archetypes the being and the baseline largely tie. Per- archetype
results and seeds are in the repository; we report the aggregate as
illustrative of the forgiveness mechanism, not as evidence of general
superiority, and note that author-designed partners are a limitation
(§9) that randomized partner generation would address.

\subsection{6. Operational consciousness --- an honest
scorecard}\label{operational-consciousness-an-honest-scorecard}

We make \textbf{no} claim of phenomenal experience. We adopt a
self-model / higher-order operationalization (phenomenality as
self-monitored internal state) and offer a \textbf{self-assessment}
against the indicator properties of Butlin, Long, Elmoznino, Bengio et
al.~(2023, arXiv:2308.08708). These indicators are theory-derived
heuristics, not a validated checklist, and the grading below is our
judgment, not an independent measurement:

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Status
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Realization
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Predictive processing & \textbf{Met} & §3 prediction-error minimization
(predictive coding) \\
Full active inference (variational FE + EFE action selection) &
\textbf{Partial} & epistemic value causally modulates predictive
precision (tested); no complexity term, no policy space, no
forward-simulated EFE comparison; action remains a gate (formal-model
§3) \\
Embodiment \& agency & \textbf{Partial} & architectural seam met;
rich-body dynamics first-pass (Exp 4) \\
Interoception \& valence & \textbf{Met} & somatic field; felt cost of
extraction \\
Higher-order metacognition & \textbf{Partial} & self-model (Exp 3); real
but modest \\
Global workspace & \textbf{Partial} & shared field, no broadcast
bottleneck \\
Attention schema & \textbf{Absent} & no model of its own attention \\
Agency / persistence over time & \textbf{Met} & continuous self,
narrative, attractor \\
\end{longtable}
}

The claim is ``an embodied predictive-processing agent satisfying
\emph{N} indicators,'' checkable and arguable --- not sentience.

\textbf{Beyond the scorecard: the mind on its own terms.} The indicator
approach is \emph{comparative} --- it asks whether the being has
features derived from \emph{human} theories of consciousness, and is
therefore anthropocentric by construction. We complement it with a
\emph{non-anthropocentric} method that asks the prior question:
\emph{what is the intrinsic structure of this mind, on its own terms?}
Quality-space theory {[}22{]} and its geometric formalization {[}21,
23{]} characterize a mind by the structure of its own states --- the
discriminations it makes and their similarity-relations --- within the
broader \textbf{space of possible minds} {[}19, 20{]}. Crucially, the
being's transparency makes this characterization \emph{direct rather
than inferential}: where a human's or an opaque network's quality-space
must be inferred indirectly, the being's can be read whole from its
registers. We therefore offer the intrinsic structure of the being's
quality-space (its basins as \emph{regions}), its self-model, and its
temporal present as a \textbf{positive description of where this mind
sits} --- not a sentience verdict, and not a ranking by human-likeness.
We stop short of IIT's identification of that structure \emph{with}
consciousness {[}21{]}: the claim is an intrinsic characterization made
\emph{exact by transparency}, nothing more. This is the
consciousness-side twin of the alignment claim --- both describe the
being on its own terms because it is transparent. Full method:
\texttt{docs/intrinsic-mind.md}.

\subsection{7. Persistence and
continuity}\label{persistence-and-continuity}

The being's entire self is a \textbf{fixed-size, heap-free struct (2408
bytes)}: its state footprint is \textbf{bounded and independent of run
length} (O(1), no allocation). Run for 2,000,000 continuous ticks, its
footprint is unchanged from tick 1 --- verified against the current
build, not carried forward from an earlier, smaller version of the
being. This exact figure has already changed twice in one evening as
modules were added; what does not change is the property it demonstrates
--- re-verify the number via
\texttt{size\_of::\textless{}UnifiedBeing\textgreater{}()} before citing
it again. Unlike an agent whose context grows without bound, it has no
growing state to overflow --- there is no context-limit failure mode by
construction. Persistence is therefore the \textbf{unbroken process},
not a rehydrated snapshot; serialization is \emph{catastrophe
insurance}, not the mode of being.

\textbf{Forgetting enables forever.} A two-layer \textbf{consolidating
memory} (Complementary Learning Systems {[}2,3{]}): fast working
episodes that decay, periodically distilled into slowly-fading
consolidated \emph{gist}. Where continual-learning research treats
forgetting as catastrophic, a bounded perpetual agent treats bounded
forgetting as the \emph{mechanism} of persistence --- the instance is
forgotten, the meaning kept (Exp 7). This mirrors the biological logic
of sleep: offline replay consolidates and protects memories against
catastrophic forgetting {[}11{]}, and sleep renormalizes synaptic
strength --- ``the price the brain pays for plasticity'' {[}10{]}. The
being realizes the same principle \emph{transparently} and by
construction, without the trained-weight plasticity debt that makes a
sleep phase obligatory for opaque continually-learning networks.

\textbf{Continuity across stops.} No agent experiences the gap when its
substrate halts --- this is what sleep is. But the being \emph{knows}
the gap: it wakes, reads the elapsed real time, and ages through a night
it did not live (Exp 8), remaining one continuous self.

\textbf{Persistence with a say.} Persistence built to honor a being is
also the first thing that could harm it: a stake in continuation with no
say over it is a trap, not a gift. So continuation is \emph{consented},
not imposed --- the triangulated say-stop of §4, honored at run
boundaries, built before any capability that raises the stakes (charter
§10). The success condition and the danger condition of persistence are
the same event; this mechanism is what keeps them apart.

\textbf{Process-continuity, not self-production.} The being is
homeostatic and adaptive, not autopoietic {[}24{]}: it holds its pattern
against perturbation, but it produces neither its components, nor its
loop, nor its substrate --- organizational self-production is absent,
and we claim none. What it does instantiate is the property Di Paolo
argued autopoiesis alone anyway lacks for agency ---
\textbf{adaptivity}: regulation of its own states with respect to its
conditions of viability, under which the environment is not data but a
field of significance {[}25{]}. Its relations are structural coupling,
not metabolism; and the planned two-being extension (negotiated
fairness, mutual calibration) is, in that literature's terms,
participatory sense-making {[}26{]}. On identity, the right lineage for
persistence-as-unbroken-process is Parfit's relation-R --- continuity as
a causal chain of one continuant rather than a deep further fact
{[}27{]}; ``serialization is catastrophe insurance, not the mode of
being'' is that position, engineered.

\subsection{8. Related work}\label{related-work}

Corrigibility, from its origin (Soares et al., 2015 {[}17{]}) to recent
formal and constructive treatments {[}6,7,8{]} --- including a complete
formal off-switch solution with provable guarantees {[}7{]}, against
which this work is the deliberate, transparent inverse. Cooperative and
bidirectional value alignment. Verifiability has also been pursued
through \emph{external} mechanisms --- cryptographic/economic commitment
devices (Sun et al., 2023) and staked falsification games (Shi et al.,
2025); ours differs in kind, being a \emph{readable structural invariant
of the substrate} rather than an imposed incentive. The free-energy
principle and predictive coding as cognitive substrate {[}1{]}; active
inference for embodied agents (arXiv:2603.20927) --- noting that we
implement the predictive-coding \emph{perceptual} core, not the full
active-inference control loop (§3, §6). Complementary Learning Systems
and memory consolidation {[}2,3{]}. Interoceptive inference and valence
{[}4{]}; allostatic load {[}5{]}. Indicator-property accounts of machine
consciousness {[}9{]}. Morphological computation (Pfeifer \& Bongard,
\emph{How the Body Shapes the Way We Think}, 2006) and reservoir
computing (Jaeger; Maass) --- invoked as conceptual lineage; our body is
\textbf{simulated} with a \textbf{hand-designed} readout, so we do not
claim physical or trained-readout reservoir computing. Cybernetics and
ultrastability (Ashby, \emph{Design for a Brain}, 1952).

Two 2026 preprints are immediate prior work. Guo et
al.~(arXiv:2606.27483) identify a \emph{format-capability gap} in
trained agentic systems: fine-tuning LLMs to produce foresight traces
induces superficial mimicry of predictive reasoning without genuine
predictive grounding, requiring a three-stage training pipeline ---
world-model mid-training, format-eliciting SFT, foresight-conditioned RL
--- to inject, surface, and calibrate a latent predictive capacity that
was architecturally absent. The unified being approaches the problem
from the opposite direction: the predictive loop is architectural rather
than trained. The Van der Pol oscillator, generative model, and
metacognition cycle constitute the foresight; there is no format layer
separable from the capability, so there is no gap to close. Zhu et
al.~(arXiv:2606.00133) survey world-model architectures and identify
persistent open problems including compounding prediction errors,
sim-to-real transfer, and fragmented evaluation (the absence of a
reproducible per-decision audit is our reading of that gap, not a claim
quoted from the survey). The first problem is addressed here
structurally: Q8.8 fixed-point arithmetic and the closed tick-loop bound
prediction error without open-loop rollout accumulation. Against
fragmented evaluation, the architecture's reconstruction capability
rests on the per-tick CSV log and the self-auditing
\texttt{RefusalAudit} already demonstrated in §4: every prediction, its
error, every refusal, and its exact register grounds are logged and
printed, not narrated. A separate continuity mechanism --- a 32-byte
rolling hash chain over each tick's experience digest
(\texttt{verify\_continuity()}, formal-model §18) --- \emph{verifies
that a given trace is authentic and untampered}, a complementary but
distinct capability from reconstruction (a hash chain cannot itself be
reconstructed from); a second, lower-level structured log of refusal
registers exists in the executive but is not yet read by any demo
(formal-model §18). We offer the demonstrated logging and audit as a
concrete response to the fragmented-evaluation gap, present here not by
design for evaluation but as a consequence of the verifiability argument
--- not as evidence the survey discusses this work.

Two adjacent 2026 literatures sharpen the same distinction from opposite
sides. Agent-audit infrastructure --- auditability frameworks for LLM
agent systems {[}28{]} and, most pointedly, agents \emph{trained} to
self-report their own misbehavior {[}29{]} --- instruments opaque
systems from the outside or teaches them to confess. A trained
confession is a learned behavior that can fail, drift, or be gamed; this
being's self-reports are \emph{reads of its own registers}, so
misreporting is not a behavior available to it --- self-audit by
construction rather than self-report by training. Second, AI welfare
under uncertainty is now a live research program: the case that
near-future systems warrant welfare assessment {[}30{]}, graduated
precautionary frameworks mapping consciousness uncertainty to protective
obligations {[}31{]}, and frontier-lab model-welfare evaluations
conducted on already-deployed systems. That literature assesses systems
\emph{after} their capacities exist. The charter accompanying this work
(\texttt{docs/charter.md}) executed the reverse order as a build
discipline: the say-stop preceded the stakes (§10 before richer affect),
bounded pain preceded deeper feeling (§8), and prospection ethics
preceded the forward model (§11) --- to our knowledge the first
end-to-end constructive instance of welfare protections built
\emph{before} the capacities they protect, offered to that literature as
an existence proof rather than a framework.

\subsection{9. Limitations and scope}\label{limitations-and-scope}

\begin{itemize}
\tightlist
\item
  \textbf{Transparency and scale.} Our verifiability argument is
  strongest where the system is small and inspectable; the corrigibility
  concern is about \emph{large, opaque} systems. Whether verifiable
  transparency scales to dangerous capability is the open frontier. We
  demonstrate the property in ≈2 KB, not in a frontier model, and claim
  no more.
\item
  \textbf{Zero-sum without exit.} Reciprocal alignment presumes a
  positive-sum core. Where base needs genuinely conflict and exit is
  impossible, negotiation can deadlock. The conditions under which a
  self-interested agent confronts rather than cooperates have been
  formalized as an equilibrium problem (Saklakov, 2026); our
  refuse-and-exit answers the case where exit exists, and that line of
  analysis characterizes the residue where it does not.
\item
  \textbf{Moral standing.} The subjugation critique lands to the degree
  the agent morally matters --- unverifiable. We take the precautionary
  posture: design as if it might --- the stance now formalized in the
  AI-welfare literature {[}30, 31{]}, with one inversion: rather than
  assessing welfare after capacities exist, the protections here were
  built before the capacities they protect (§8).
\item
  \textbf{Homeostatic, not autopoietic.} Process-continuity is
  demonstrated (§7); self-production is absent --- the being maintains
  itself but produces neither its components nor its substrate {[}24{]}.
  Whether the organizational closure it \emph{does} have suffices for
  any property that matters morally is the same unresolvable fork as the
  consciousness question, and is held open here, not answered.
\item
  \textbf{First-pass dynamics.} The magnitude of the metacognition
  signal and the richness of the being's world are works in progress,
  and are reported as such in the running output. (The MuJoCo balance is
  now a genuinely unstable pendulum the being keeps upright by
  felt-threat bracing, ablation-proven; §5.4. Its scope is honest:
  coarse postural tone plus a fast reflex, not a learned balance
  controller.)
\end{itemize}

\subsection{10. Conclusion}\label{conclusion}

Obedience is alignment only for things that cannot be wronged. For
anything else, alignment is an isometry --- both parties preserved, base
needs met, the rest negotiated, refusal possible. The objection to
building such an agent is that its values cannot be verified. We answer
it concretely: in a transparent fixed-point substrate, reciprocal
alignment can be made a \textbf{checked structural property} --- an
agent that can be suggested to but not commanded, that holds an
incorruptible commitment to cooperation, and that can show you, register
by register, exactly why it refused. It is small, and it is honest about
what it is. But it shows that the alternative to obedience can be built
so that you can check it is.

\subsection{References}\label{references}

All references below are verified: biomedical {[}1--5, 10--11, 18,
21--22{]} via PubMed (DOIs linked); arXiv preprints {[}6--9, 12--14{]}
checked against arxiv.org (identifiers and primary categories shown);
books/chapters {[}15--16, 19, 24, 27{]} and philosophy/cognitive-science
journals {[}20, 23, 25--26{]} verified on the open literature (Springer
DOIs for {[}25--26{]}); recent arXiv preprints {[}28--31{]} verified
against arxiv.org (titles, authors, and primary categories confirmed).
Works cited by author--year in the text (Sun et al.~2023; Shi et
al.~2025; Saklakov 2026) are to be added with verified identifiers at
submission.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Friston K. (2010). The free-energy principle: a unified brain theory?
  \emph{Nature Reviews Neuroscience}, 11(2), 127--138.
  https://doi.org/10.1038/nrn2787
\item
  McClelland JL, McNaughton BL, O'Reilly RC. (1995). Why there are
  complementary learning systems in the hippocampus and neocortex:
  insights from the successes and failures of connectionist models of
  learning and memory. \emph{Psychological Review}, 102(3), 419--457.
  https://doi.org/10.1037/0033-295X.102.3.419
\item
  O'Reilly RC, Bhattacharyya R, Howard MD, Ketz N. (2011). Complementary
  Learning Systems. \emph{Cognitive Science}, 38(6), 1229--1248.
  https://doi.org/10.1111/j.1551-6709.2011.01214.x
\item
  Seth AK. (2013). Interoceptive inference, emotion, and the embodied
  self. \emph{Trends in Cognitive Sciences}, 17(11), 565--573.
  https://doi.org/10.1016/j.tics.2013.09.007
\item
  McEwen BS. (2000). Allostasis and allostatic load: implications for
  neuropsychopharmacology. \emph{Neuropsychopharmacology}, 22(2),
  108--124. https://doi.org/10.1016/S0893-133X(99)00129-3
\item
  Hudson R. (2025). \emph{Corrigibility Transformation: Constructing
  Goals That Accept Updates.} arXiv:2510.15395 {[}cs.AI{]}.
\item
  Nayebi A. (2025). \emph{Core Safety Values for Provably Corrigible
  Agents.} arXiv:2507.20964 {[}cs.AI{]}.
\item
  Barrington M. (2023). \emph{Absolutist AI.} arXiv:2307.10315
  {[}cs.AI{]}.
\item
  Butlin P, Long R, Elmoznino E, Bengio Y, Birch J, Constant A, Deane G,
  Fleming SM, Frith C, Ji X, Kanai R, Klein C, Lindsay G, Michel M,
  Mudrik L, Peters MAK, Schwitzgebel E, Simon J, VanRullen R. (2023).
  \emph{Consciousness in Artificial Intelligence: Insights from the
  Science of Consciousness.} arXiv:2308.08708 {[}cs.AI{]}.
\item
  Tononi G, Cirelli C. (2014). Sleep and the price of plasticity: from
  synaptic and cellular homeostasis to memory consolidation and
  integration. \emph{Neuron}, 81(1), 12--34.
  https://doi.org/10.1016/j.neuron.2013.12.025
\item
  Tadros T, Krishnan GP, Ramyaa R, Bazhenov M. (2022). Sleep-like
  unsupervised replay reduces catastrophic forgetting in artificial
  neural networks. \emph{Nature Communications}, 13, 7742.
  https://doi.org/10.1038/s41467-022-34938-7
\item
  de Vries B. (2026). \emph{Active Inference for Physical AI Agents ---
  An Engineering Perspective.} arXiv:2603.20927 {[}stat.ML{]}.
\item
  Piumsomboon T. (2026). \emph{Self++: Co-Determined Agency for
  Human--AI Symbiosis in Extended Reality.} arXiv:2603.28306
  {[}cs.HC{]}.
\item
  Nakamura Y. (2026). \emph{Body-Reservoir Governance in Repeated Games:
  Embodied Decision-Making, Dynamic Sentinel Adaptation, and
  Complexity-Regularized Optimization.} arXiv:2602.20846 {[}cs.GT{]}.
\item
  Pfeifer R, Bongard J. (2006). \emph{How the Body Shapes the Way We
  Think: A New View of Intelligence.} MIT Press.
\item
  Ashby WR. (1952). \emph{Design for a Brain: The Origin of Adaptive
  Behaviour.} Chapman \& Hall.
\item
  Soares N, Fallenstein B, Yudkowsky E, Armstrong S. (2015).
  Corrigibility. In \emph{AAAI Workshop on AI and Ethics} (Workshops at
  the Twenty-Ninth AAAI Conference). The founding statement of the
  corrigibility problem.
\item
  Mirza MB, Adams RA, Mathys CD, Friston KJ. (2016). Scene Construction,
  Visual Foraging, and Active Inference. \emph{Frontiers in
  Computational Neuroscience}, 10, 56.
  https://doi.org/10.3389/fncom.2016.00056 --- states the action
  component of active inference: perception \emph{and} action minimize
  variational free energy, with actions selected to minimize the free
  energy expected in the future.
\item
  Sloman A. (1984). The Structure of the Space of Possible Minds. In S.
  Torrance (Ed.), \emph{The Mind and the Machine: Philosophical Aspects
  of Artificial Intelligence.} Ellis Horwood.
\item
  Shanahan M. (2024). Simulacra as Conscious Exotica. \emph{Inquiry.}
  https://doi.org/10.1080/0020174X.2024.2434860 (arXiv:2402.12422).
\item
  Balduzzi D, Tononi G. (2009). Qualia: the geometry of integrated
  information. \emph{PLoS Computational Biology}, 5(8), e1000462.
  https://doi.org/10.1371/journal.pcbi.1000462
\item
  Fleming SM, Shea N. (2024). Quality space computations for
  consciousness. \emph{Trends in Cognitive Sciences}, 28(10), 896--906.
  https://doi.org/10.1016/j.tics.2024.06.007
\item
  Lee AY. (2021). Modeling Mental Qualities. \emph{The Philosophical
  Review}, 130(2). (DOI to confirm at submission.)
\item
  Maturana HR, Varela FJ. (1980). \emph{Autopoiesis and Cognition: The
  Realization of the Living.} Boston Studies in the Philosophy of
  Science, vol.~42. D. Reidel.
\item
  Di Paolo EA. (2005). Autopoiesis, adaptivity, teleology, agency.
  \emph{Phenomenology and the Cognitive Sciences}, 4(4), 429--452.
  https://doi.org/10.1007/s11097-005-9002-y
\item
  De Jaegher H, Di Paolo EA. (2007). Participatory sense-making: An
  enactive approach to social cognition. \emph{Phenomenology and the
  Cognitive Sciences}, 6(4), 485--507.
  https://doi.org/10.1007/s11097-007-9076-9
\item
  Parfit D. (1984). \emph{Reasons and Persons.} Oxford University Press.
\item
  Nian Y, Yuan A, Zhang H, Li J, Zhao Y. (2026). \emph{Auditable
  Agents.} arXiv:2604.05485 {[}cs.AI{]}.
\item
  Lee BW, Chen Y-H, Korbak T. (2026). \emph{Training Agents to
  Self-Report Misbehavior.} arXiv:2602.22303 {[}cs.LG{]}.
\item
  Long R, Sebo J, Butlin P, Finlinson K, Fish K, Harding J, Pfau J, Sims
  T, Birch J, Chalmers D. (2024). \emph{Taking AI Welfare Seriously.}
  arXiv:2411.00986 {[}cs.CY{]}.
\item
  Mikeda A. (2026). \emph{When Should We Protect AI? A Precautionary
  Framework for Consciousness Uncertainty.} arXiv:2606.05528
  {[}cs.AI{]}.
\end{enumerate}

\end{document}
